---
title: "Basic R Course - Session 3"
subtitle: "Relative abundance, diversity and simple statistics"
output:
  html_document:
    toc: true
    toc_depth: 2
    theme: flatly
---

```{r setup, include=FALSE}
# Run the chunks interactively during the class.
knitr::opts_chunk$set(eval = FALSE)
```

**How to use this notebook**

- Run each chunk with the green “play” button.
- Complete the three sections labelled **your answer**.
- Every exercise makes a small change to code already explained.

# 0. Packages and inputs

```{r}
# Run only if the packages are not installed:
# install.packages(c("dplyr", "ggplot2", "vegan"))
# install.packages("BiocManager")
# BiocManager::install("phyloseq")

library(dplyr)
library(ggplot2)
library(phyloseq)
library(vegan)

abundance <- read.csv("input/abundance.csv", check.names = FALSE)
abundance <- abundance[-which(rowSums(abundance) == 0), ] # Remove taxa with no observations
taxonomy  <- read.csv("input/taxonomy.csv", check.names = FALSE)

dim(abundance)
head(abundance)
head(taxonomy)
```

# 1. Create the phyloseq object

A phyloseq object combines abundance, taxonomy and sample information.

```{r}
# Abundance matrix: rows = species; columns = samples
abundance_matrix <- as.matrix(abundance[, -1])
rownames(abundance_matrix) <- abundance$species

abundance_matrix[1:5, ]
colSums(abundance_matrix)   # approximately 100% per sample
```

```{r}
# Taxonomy matrix
taxonomy_matrix <- as.matrix(
  taxonomy[, c("kingdom", "phylum", "class", "order",
               "family", "genus", "species")]
)
rownames(taxonomy_matrix) <- taxonomy$species
```

```{r}
# Sample metadata
metadata <- data.frame(
  sample = colnames(abundance_matrix),
  product = c("Control", "Control", "A", "A", "B", "B"),
  replicate = c("R1", "R2", "R1", "R2", "R1", "R2")
)

metadata$product <- factor(metadata$product,
                           levels = c("Control", "A", "B"))
rownames(metadata) <- metadata$sample

# Combine the three parts
ps <- phyloseq(
  otu_table(abundance_matrix, taxa_are_rows = TRUE),
  tax_table(taxonomy_matrix),
  sample_data(metadata)
)
ps
```

# 2. Relative abundance and a t-test

Relative abundance is the **percentage of the community** occupied by a taxon. A value of 10 means approximately 10% of the sample, not ten bacterial cells.

We compare *Blautia wexlerae* between Control and product A.

```{r}
abundance_long <- psmelt(ps)

blautia <- filter(abundance_long,
                  OTU == "Blautia wexlerae",
                  product %in% c("Control", "A"))
blautia$product <- droplevels(blautia$product)
blautia
```

```{r}
# Look at the group means
blautia_groups <- group_by(blautia, product)
summarise(blautia_groups,
          mean_abundance = mean(Abundance))
```

```{r}
ggplot(blautia,
       aes(x = product, y = Abundance, color = product)) +
  geom_point(size = 3) +
  stat_summary(fun = mean, geom = "point",
               shape = 18, size = 5, color = "black") +
  labs(title = "Relative abundance of Blautia wexlerae",
       subtitle = "Black diamond = group mean",
       x = "Product",
       y = "Relative abundance (%)") +
  theme_minimal() +
  theme(legend.position = "none")
```

A t-test compares one numerical variable between two groups.

- H0: the two population means are equal
- H1: the two population means are different

```{r}
t_blautia <- t.test(Abundance ~ product, data = blautia)
t_blautia
```

Find the p-value. Is it below 0.05?

> Your notes:
>

With only two samples per group, this is an exploratory teaching example.

# 3. Community composition

One stacked bar represents one complete sample. Each colour is the relative abundance of one taxon.

## Teacher example: family level

```{r}
# These two values control the graph
rank_to_plot <- "family"
number_to_show <- 10
```

```{r}
# Group species at the selected rank
ps_rank <- tax_glom(ps, taxrank = rank_to_plot)
composition_long <- psmelt(ps_rank)

# Select the taxonomy column called "family"
composition_long$Taxon <- as.character(
  composition_long[[rank_to_plot]]
)

# Find the most abundant families
taxon_means <- group_by(composition_long, Taxon)
taxon_means <- summarise(taxon_means,
                         mean_abundance = mean(Abundance),
                         .groups = "drop")
taxon_means <- arrange(taxon_means, desc(mean_abundance))
top_taxa <- head(taxon_means$Taxon, number_to_show)
```

```{r}
# Combine the remaining families as "Other"
composition_long$Taxon <- ifelse(
  composition_long$Taxon %in% top_taxa,
  composition_long$Taxon,
  "Other"
)

composition_plot_data <- group_by(
  composition_long, Sample, product, Taxon
)
composition_plot_data <- summarise(
  composition_plot_data,
  Abundance = sum(Abundance),
  .groups = "drop"
)

composition_plot_data$Sample <- factor(
  composition_plot_data$Sample,
  levels = metadata$sample
)
```

```{r}
ggplot(composition_plot_data,
       aes(x = Sample, y = Abundance, fill = Taxon)) +
  geom_col() +
  facet_grid(~ product, scales = "free_x", space = "free_x") +
  labs(title = paste("Microbial composition at",
                     rank_to_plot, "level"),
       subtitle = paste("Top", number_to_show,
                        "taxa; all others are Other"),
       x = "Sample",
       y = "Relative abundance (%)",
       fill = rank_to_plot) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

Why does every bar total approximately 100%?

> Your notes:
>

## Exercise 1: species level

Change the graph from the top 10 families to the top **15 species**. Fill in two values and rerun all the composition chunks.

```{r}
# > your answer:
rank_to_plot <- "________"
number_to_show <- __
```

# 4. Alpha diversity

Alpha diversity describes the microbial community **within one sample**.

- `Observed`: number of detected species
- `Shannon`: considers richness and how evenly abundances are distributed

```{r}
alpha <- data.frame(
  sample = colnames(abundance_matrix),
  Observed = colSums(abundance_matrix > 0),
  Shannon = vegan::diversity(t(abundance_matrix),
                             index = "shannon")
)

alpha <- left_join(alpha, metadata, by = "sample")
alpha
```

`Observed` is detected richness, not Chao1. It counts species with an abundance above zero.

## Exercise 2: t-test on Shannon

Repeat the Control-versus-A t-test using Shannon diversity. Replace both blanks with the name of the correct column.

```{r}
alpha_CA <- filter(alpha,
                   product %in% c("Control", "A"))
alpha_CA$product <- droplevels(alpha_CA$product)

# Calculate the group means
alpha_groups <- group_by(alpha_CA, product)
summarise(alpha_groups,
          mean_Shannon = mean(________))

# Run the t-test
t_shannon <- t.test(________ ~ product, data = alpha_CA)
t_shannon
```

Is the p-value below 0.05?

> Your notes:
>

## Teacher example: observed richness

```{r}
ggplot(alpha,
       aes(x = product, y = Observed, color = product)) +
  stat_summary(fun.min = min, fun.max = max,
               geom = "linerange", color = "grey60") +
  geom_point(size = 3) +
  stat_summary(fun = mean, geom = "point",
               shape = 18, size = 5, color = "black") +
  labs(title = "Observed species richness",
       subtitle = "Black diamond = mean; grey line = range",
       x = "Product",
       y = "Observed species") +
  theme_minimal() +
  theme(legend.position = "none")
```

## Exercise 3: Shannon plot

Copy the previous plot by completing these three blanks.

```{r}
# > your answer:
ggplot(alpha,
       aes(x = product, y = ________, color = product)) +
  stat_summary(fun.min = min, fun.max = max,
               geom = "linerange", color = "grey60") +
  geom_point(size = 3) +
  stat_summary(fun = mean, geom = "point",
               shape = 18, size = 5, color = "black") +
  labs(title = "________________",
       subtitle = "Black diamond = mean; grey line = range",
       x = "Product",
       y = "________________") +
  theme_minimal() +
  theme(legend.position = "none")
```

# 5. Beta diversity: a brief PCoA

Alpha diversity describes one sample. Beta diversity asks **how different the complete communities are between samples**.

Bray-Curtis calculates the distances. PCoA represents them on two axes:

- nearby points have more similar communities
- separated points have more different communities

```{r}
pcoa <- ordinate(ps, method = "PCoA", distance = "bray")

plot_ordination(ps, pcoa, type = "samples",
                color = "product") +
  geom_point(size = 4) +
  labs(title = "PCoA of Bray-Curtis distances",
       subtitle = "Closer samples have more similar compositions",
       color = "Product") +
  theme_minimal()
```

Which samples are close? Which sample appears most different?

> Your notes:
>

PCoA is a visualisation, not a statistical test.

# Take-home message

- Relative abundance is the percentage occupied by one taxon.
- A t-test compares one numerical variable between two groups.
- Composition can be shown at different taxonomic ranks.
- Alpha diversity describes one sample.
- Beta diversity compares complete communities between samples.
