---
title: "Basic R Course.- Handling tables with dplyr"
output: html_document
---



```{r}
# install.packages("dplyr")    # only the first time

library(dplyr)

acids <- read.csv("input/acids.csv")

acids
dim(acids)     # 6 rows (the samples) and 9 columns
```

# 1. The 6 dplyr verbs

- `filter()` picks ROWS
- `select()` picks COLUMNS
- `arrange()` sorts the rows
- `mutate()` adds a new column
- `summarise()` reduces to a single number
- `group_by()` does the above per group

They are all written the same way, and the first argument is ALWAYS the table: `verb(table, what_I_want)`

## filter() -> picks ROWS

You give it a condition and it keeps the rows that meet it. The columns are untouched.

```{r}
# "give me the rows whose product is B"
filter(acids, product == "B")
#   from 6 rows -> 2 remain

# "give me the rows with more than 1000 ug/ml of butyric acid"
filter(acids, butyric > 1000)
#   from 6 rows -> 3 remain
```

CAREFUL: to compare you write a DOUBLE equals sign `==`

### Exercise 1a

Keep only the Control rows.


```{r}
# your answer here
```

### Exercise 1b

Now a filter with a number instead of a word: keep only the rows whose acetic acid is above 3000.


```{r}
# your answer here
```

## select() -> picks COLUMNS

The opposite of `filter()`. All rows are kept.

```{r}
# "give me only these two columns"
select(acids, sample, butyric)
#   still 6 rows, but out of 9 columns only 2 remain

# A minus sign REMOVES a column
select(acids, -formic)
```

### Exercise 2a

Keep only the columns sample, product and acetic.


```{r}
# your answer here
```

### Exercise 2b

Now the other way round: show every column EXCEPT propionic (use the minus sign).


```{r}
# your answer here
```

## arrange() -> sorts the rows

It removes nothing and adds nothing, it only changes order.

```{r}
# sort the samples by butyric acid
arrange(acids, butyric)          # from lowest to highest
arrange(acids, desc(butyric))    # desc = highest to lowest
```

### Exercise 3a

Sort the samples from highest to lowest propionic acid.


```{r}
# your answer here
```

### Exercise 3b

Now without `desc()`: sort the samples from lowest to highest formic acid.


```{r}
# your answer here
```

## mutate() -> adds a new column

Computes something from the existing columns and sticks it on at the end.

The "good" fatty acids are butyric and propionic:

```{r}
# add the two "good" acids into a single new column
mutate(acids, scfa = butyric + propionic)
#   an scfa column appears at the end: 9 columns -> 10
#   (scfa = short chain fatty acids)
```

IMPORTANT: `mutate()` only shows it on screen. To keep it in the table you have to store it with `<-`:

```{r}
acids <- mutate(acids, scfa = butyric + propionic)
acids
```

### Exercise 4a

Create a column `ratio` that is butyric divided by acetic, rounded to 2 decimals.


```{r}
# your answer here
```

### Exercise 4b

Now an easier one, just a sum: create a column `acetic_lactic` that adds acetic and lactic.


```{r}
# your answer here
```
### Exercise 4.1

Two separate lines:

a) sort the samples from highest to lowest scfa
b) show only the columns sample and scfa


```{r}
# your answer here
```

## summarise() -> reduces to a single number

It condenses all the rows into one.

In session 1 we would have written this:

```{r}
mean(acids$butyric)
```

`summarise()` does exactly the same thing, but hands it back as a TABLE, with a column name that we choose:

```{r}
# the average butyric acid across all 6 samples, as a table
summarise(acids, mean_butyric = mean(butyric))
#   from 6 rows -> 1 row
```

So why bother, if `mean()` already worked? Because of what comes next: `summarise()` can do this per group, and plain `mean()` cannot.

You can ask for several numbers at once:

```{r}
summarise(acids,
          mean_butyric   = mean(butyric),
          mean_propionic = mean(propionic))
```

### Exercise 5a

Compute the mean and the maximum of propionic acid.


```{r}
# your answer here
```

### Exercise 5b

The same idea with two different summaries and another acid: compute the minimum and the mean of formic acid.


```{r}
# your answer here
```

## group_by() -> doing it per group

Everything above gives ONE number for the whole table. But what we actually want is one number for the Control, one for A and one for B, so that we can compare them.

**First, how we would do it BY HAND.** Pick out one product, compute its mean. Then the next one.

```{r}
control_rows <- filter(acids, product == "Control")
mean(control_rows$butyric)          # 491.8

a_rows <- filter(acids, product == "A")
mean(a_rows$butyric)                # 1237.5

b_rows <- filter(acids, product == "B")
mean(b_rows$butyric)                # 2319.8
```

It works. But it took 6 lines for 3 products, the results come out loose instead of in a table, and with 20 products it would be unbearable.

**Now the same thing with group_by():**

```{r}
# average butyric acid per product, in one table
by_product <- group_by(acids, product)
summarise(by_product, mean_butyric = mean(butyric))
```

Same three numbers, in one table, in two lines.

`group_by()` on its own displays nothing new: all it does is mark which rows belong together. The work is done by the `summarise()` that comes after it, which now runs once per group instead of once for the whole table.

### Exercise 6a

Now the same thing, but with PROPIONIC acid: compute its mean per product. Two lines: first group, then summarise.


```{r}
# your answer here
```

### Exercise 6b
A different question, still per product: how MANY samples does each product have? (use n(), which counts rows)

```{r}
# your answer here
```

# 2. Chaining steps

One verb is almost never enough. To do several things in a row you store each step in a variable with `<-` and pass it to the next one. It is like an assembly line.

```{r}
# step 1: keep only replicate 1's rows
r1_only <- filter(acids, replicate == 1)   # 6 rows -> 3
# step 2: from that result, keep three columns
small   <- select(r1_only, sample, formic, acetic)
small
```

The result of step 1 is the input table of step 2. Give them names that say what they contain, not step1/step2.

## The pipe (optional, just so you recognise it)

On the internet you will constantly see the symbol `|>` (or the older `%>%`), the "pipe". It chains the same steps without inventing a name for each one: it takes the table on the left and hands it to the next verb as its first argument.

```{r}
# exactly the same as the two lines above, but in one line
acids |> filter(replicate == 1) |> select(sample, formic, acetic)
```

You do NOT need the pipe for this course. We show it once, actually used, so that it does not startle you when you meet it online.

## Exercise 8

Keep the rows of product B and sort them from highest to lowest butyric acid, storing each step in a variable.


```{r}
# your answer here
```

## Exercise 8b (optional: with the pipe)

Redo Exercise 8 in a single line using the pipe `|>`, without storing anything in between.


```{r}
# your answer here
```

# 3. The summary table of the experiment

Three verbs chained: group, summarise and sort.

```{r}
# 1) group the rows by product
by_product <- group_by(acids, product)

# 2) one mean per group, for the two acids we care about
summary_table <- summarise(by_product,
                           butyric   = round(mean(butyric), 1),
                           propionic = round(mean(propionic), 1))
summary_table

# 3) sort it, biggest first
summary_table <- arrange(summary_table, desc(butyric))
summary_table
```

Product B produces far more butyric and propionic acid than the Control.

```{r}
write.csv(summary_table, "output/acids_summary.csv", row.names = FALSE)
```

## Exercise 9

Chaining two verbs, storing each step:

a) keep the rows of replicate 2
b) sort them from highest to lowest propionic acid
c) save the result in `output/replicate2_propionic.csv`


```{r}
# your answer here
```

# 4. Do the two replicates agree?

Each condition was fermented twice, separately (R1 and R2). Before summarising anything you have to check whether the two give the same answer. If they do, the experiment is reproducible.

We will compare ALL SIX acids, not just one. If we only looked at one, the answer would depend on which one we happened to pick.

**Step 1:** we split the two replicates into two tables.

```{r}
r1 <- filter(acids, replicate == 1)     # Control_R1, A_R1, B_R1
r2 <- filter(acids, replicate == 2)     # Control_R2, A_R2, B_R2

r1
r2
```

**Step 2:** we keep only the six acid columns of each table.

```{r}
values_r1 <- select(r1, formic:citric)
values_r2 <- select(r2, formic:citric)

values_r1
values_r2
```

**Step 3:** we compare them, one idea per line.

When you subtract two tables of the same size, R subtracts them cell by cell: the formic of Control_R1 minus the formic of Control_R2, and so on for all 18 numbers.

```{r}
gap <- abs(values_r1 - values_r2)     # HOW FAR APART the two are
gap
```

Why `abs()`? Without it we would get negative numbers whenever R2 is the bigger one. The minus sign only tells us WHICH one was bigger, and we do not care about that: we want to know how far apart they are. `abs()` drops the sign and keeps the size. Try it without `abs()` and compare.

```{r}
average <- (values_r1 + values_r2) / 2   # the TYPICAL VALUE there
average
```

CAREFUL: this is not the mean of an acid across samples. It is the middle point between the two replicates of that one acid in that one condition.

```{r}
diff_pct <- round(100 * gap / average, 1)   # gap, as a % of the typical value
diff_pct
```

**WHY DO WE DIVIDE?** Two real examples from this table:

```
                     R1       R2      gap   typical value    %
  butyric in A     1636      839      796         1238      64
  formic  in B    22334    26137     3804        24236      16
```

If we stopped at the gap, the formic acid of B (3804) would look FIVE TIMES WORSE than the butyric acid of A (796). It is the other way round: 3804 out of 24000 is small change, while 796 out of 1238 is enormous.

Think of two people measuring something and disagreeing by 1 cm. On a 2 m table that is nothing; on a 3 cm screw it is a disaster. Same centimetre. That is why the gap only means something once you compare it with the size of what you are measuring.

Dividing also makes the six acids comparable with each other, which is what lets us read the whole table at once.

And we glue the product names back on, to read it easily:

```{r}
comparison <- data.frame(product = r1$product, diff_pct)
comparison
```

```
product formic acetic propionic lactic butyric citric
Control    2.7    4.9      60.1    4.6    37.9    2.9
      A    0.7   11.7      43.6   29.2    64.3   32.9
      B   15.7    6.7      25.1    3.7     5.5   14.6
```

## Exercise 10

This one has no code: it is about READING the table above. Remember what each number means: "the two replicates differ by this percentage of the normal value". Small = they agree.

a) Look at the three rows. Which product has the most reproducible replicates, that is, the smallest numbers?
b) Now look at the six columns. Which TWO acids give the biggest numbers in all three products?

**CONCLUSION FOR THE CLASS:**

1. Product B is the most reproducible condition, product A the least. Now we can say this looking at a whole row, not at a single number.
2. Formic, acetic and lactic acid repeat almost exactly (under 12% nearly everywhere). Propionic and butyric are the noisy ones, in every condition.
3. And here is the uncomfortable part: propionic and butyric are precisely the two acids that matter for our hypothesis, the ones linked to insulin. The measurements we care about most are the least reproducible ones. That is exactly why the authors of the original paper reported these acids descriptively, without statistics.

The lesson: reproducibility is not the same in every condition or for every measurement, and you have to LOOK at it rather than assume it.

---

# END OF SESSION 2 (dplyr)

- `filter()` picks ROWS (trims vertically)
- `select()` picks COLUMNS (trims horizontally)
- `arrange()` sorts
- `mutate()` adds a column
- `summarise()` reduces
- `group_by()` makes summarise work per group

Chain steps by storing each result in a variable. Check the replicates BEFORE summarising them.

(You have also met the pipe `|>` once, in section 2. You do not need it for this course; we just wanted it not to startle you when you see it online.)
