---
title: "Basic R Course - Session 1 - Student Copy"
output: html_document
---

# BASIC R COURSE  -  SESSION 1        *** STUDENT COPY ***
## First steps: variables, vectors, classes and data frames

 How to use this file:
   - Follow along while we go through it together.
   - Put the cursor on a line and press Ctrl + Enter
     (Cmd + Enter on Mac) to run that line.
   - Anything starting with # is a comment: R ignores it.
   - Where you see  "# > your answer:"  the code is yours to
     write. There is space left underneath for it.
   - Save this file in your working folder. It is your
     notes for the day.

## 0. BEFORE WE START: THE WORKING DIRECTORY

 R is always "standing" in some folder. When you ask it for a
 file, it looks there. This is the number one problem for
 beginners.

 What to do, once, at the start:
   1) Decide your working directory (where you want to work)
   2) Open RStudio and create a script:
      File -> New File -> R Script, and save it in that folder
   3) RStudio menu: Session -> Set Working Directory ->
      To Source File Location

```{r}
getwd()          # which folder am I in?
```

```{r}
list.files()     # what is in this folder?
```

### Changing folder by hand
Replace the path with the one on YOUR computer:

```{r}
# setwd("C:/Users/your_user/.../curso_R_datos")
```

CAREFUL ON WINDOWS: copying a path gives you backslashes \
  WRONG  "C:\Users\alba\curso_R_datos"
  RIGHT  "C:/Users/alba/curso_R_datos"
  RIGHT  "C:\\Users\\alba\\curso_R_datos"

## 1. R IS A CALCULATOR

```{r}
2 + 2
10 / 4
3 ^ 2          # power
sqrt(16)       # square root
```

## 2. STORING THINGS IN VARIABLES
   The arrow  <-  means "store this here".
   Keyboard shortcut: Alt + minus sign

```{r}
weight <- 70
height <- 1.75
weight          # typing the name shows its content
```

### MINI EXERCISE
Compute the body mass index (weight divided by height squared)
and store it in a variable called  bmi
  you need:  the ^ operator  and  <-

> your answer:

```{r}

```

Names are case sensitive: Weight and weight are different.
Use short names, without spaces or accents.

## 3. VECTORS: several values in a single variable
   Built with c()  (for "combine")

```{r}
numbers <- c(1, 2, 3, 4, 5)
numbers
```

```{r}
names_list <- c("Ana", "Luis", "Marta")
names_list
```

A very handy shortcut: a colon generates a sequence

```{r}
1:5
10:15
```

How many elements does a vector have

```{r}
length(numbers)
```

## 4. POSITIONS: pulling out specific elements
   Square brackets [ ] are used. IMPORTANT: in R counting
   starts at 1, not at 0.

```{r}
numbers <- 1:5
numbers
```

```{r}
numbers[1]        # the first one
numbers[3]        # the third one
numbers[2:4]      # from the second to the fourth
numbers[c(1, 5)]  # the first and the fifth
numbers[-1]       # ALL BUT the first
```

You can also select by condition

```{r}
numbers[numbers > 3]
```

## 5. OPERATIONS ON THE WHOLE VECTOR AT ONCE
   This is one of the great advantages of R.

```{r}
numbers
numbers + 5        # adds 5 to EACH value
numbers * 2
```

Summary functions

```{r}
mean(numbers)      # mean
sd(numbers)        # standard deviation
min(numbers)
max(numbers)
sum(numbers)
round(mean(numbers), 1)   # round to 1 decimal
```

```{r}
summary(numbers)   # the whole summary in one go
```

### EXERCISE 1
Create a vector called  marks  with the values 7, 8, 5, 9, 6.
Then compute its mean rounded to 2 decimals, and its
standard deviation.

  you need:  c()   mean()   round()   sd()

> your answer:

```{r}

```

## 6. DATA TYPES: the class() function
   Knowing the type of a column prevents 90% of the errors.

```{r}
class(numbers)      # "numeric"   -> numbers
class(names_list)   # "character" -> text
class(TRUE)         # "logical"   -> true / false
```

Comparisons: questions with a yes/no answer

```{r}
numbers > 3         # gives TRUE / FALSE for each element
sum(numbers > 3)    # TRUE counts as 1: how many are above 3
```

A typical mistake: if you mix text and numbers, EVERYTHING
becomes text

```{r}
mixed <- c(1, 2, "three")
mixed
class(mixed)            # "character"  <- careful!
```

## 7. COUNTING CATEGORIES: the table() function

```{r}
sex   <- c("M", "F", "F", "M", "F")
group <- c("control", "treated", "control", "treated", "control")
```

```{r}
table(sex)
table(group)
```

table() also cross-tabulates two variables

```{r}
table(sex, group)
```

### EXERCISE 2
Create a vector  days  with
"monday","tuesday","monday","monday","tuesday","wednesday"
and count how many times each day appears.

  you need:  c()   table()

> your answer:

```{r}

```

## 8. DATA FRAMES: the table of R
   This is what we will use from now on. Each column is a
   vector, and they must all have the same length.

```{r}
patients <- data.frame(
  name  = c("Ana", "Luis", "Marta", "Jorge"),
  age   = c(34, 45, 29, 52),
  group = c("control", "treated", "control", "treated")
)
```

```{r}
patients
```

### Exploring a data frame: ALWAYS the first thing to do

```{r}
nrow(patients)     # number of ROWS    (= samples)
ncol(patients)     # number of COLUMNS (= variables)
dim(patients)      # rows and columns in one go
names(patients)    # the column names
str(patients)      # structure: the type of each column
head(patients, 2)  # the first 2 rows
tail(patients, 2)  # the last 2 rows
summary(patients)  # a summary of each column
```

### Accessing a column: the $ symbol

```{r}
patients$age
mean(patients$age)
table(patients$group)
```

### Accessing by position: [row, column]

```{r}
patients[1, ]        # the whole row 1
patients[, 2]        # the whole column 2
patients[1, 2]       # row 1, column 2
patients[1:2, ]      # the first two rows
```

### Filtering rows by a condition

```{r}
patients[patients$age > 40, ]
patients[patients$group == "control", ]
```

  Careful: to compare you use a DOUBLE equals sign  ==
  A single  =  means something else.

### Creating a new column

```{r}
patients$age_months <- patients$age * 12
patients
```

### EXERCISE 3
Using the data frame  patients:
  a) how many rows and columns does it have
  b) the mean age
  c) keep only the rows of the "treated" group

  you need:  dim()   mean()   $   [ , ]   ==

> your answer:

```{r}

```

### EXERCISE 4 (recap)
Create a data frame  plants  with:
  pot:     1, 2, 3, 4, 5, 6
  water:   "low","low","low","high","high","high"
  height:  12, 15, 11, 22, 25, 20
Then:
  a) look at its structure
  b) how many pots there are of each watering level
  c) the mean height of the pots with "high" watering

  you need:  data.frame()   str()   table()   mean()   $   [ ]

> your answer:

```{r}

```

## 9. ASKING FOR HELP

```{r}
?mean            # opens the help page of a function
help(round)
```

And if you do not even know the name of the function, search
the internet for "how to ... in R". That is what everybody
does, always.

## PART 2 OF THE SESSION: OUR OWN DATA

## 10. THE EXPERIMENT WE ARE GOING TO ANALYSE

 Two food products (dairy fermentates, in powder form) were
 tested to see whether they could help control blood glucose.
 Each product went through three stages:

   1) Artificial DIGESTION in the lab (mouth, stomach,
      intestine), to simulate what would happen if eaten.
   2) FERMENTATION with human gut bacteria for 24 h. The
      bacteria transform the product and produce organic
      acids, mainly SHORT CHAIN FATTY ACIDS (butyric,
      propionic, acetic).
   3) The resulting liquid is put in contact with human
      PANCREATIC CELLS and the INSULIN they release is
      measured.

 HYPOTHESIS: if a product makes the bacteria produce more
 butyric and propionic acid, the cells should release more
 insulin.

 CONDITIONS (3):
   Control  ->  the bacteria with no product added
   A        ->  product A (high in lactose, low in protein)
   B        ->  product B (very high in protein, no lactose)

 REPLICATES (2 per condition): each condition was fermented
 in TWO independent vessels, R1 and R2. These are BIOLOGICAL
 REPLICATES: two real repetitions of the experiment.

   Control_R1   Control_R2
   A_R1         A_R2
   B_R1         B_R2      -> 6 samples in total

 MEASUREMENTS. Careful, because they are not all taken at
 the same stage of the pipeline:

   acids and bacteria  ->  measured ONCE in each vessel.
                           6 vessels = 6 measurements.

   insulin             ->  the liquid from each vessel was
                           put on the beta cells on 3
                           SEPARATE OCCASIONS, using a
                           different batch of cells each
                           time (column "experiment").
                           6 vessels x 3 = 18 measurements.

 So the 3 experiments are NOT 3 new fermentations: they are
 the SAME fermented liquid measured three times. They tell us
 how repeatable the CELL ASSAY is, not how repeatable the
 BIOLOGY is. Telling those two things apart is one of the
 goals of this course.

## 11. READING A CSV FILE

 A CSV is a text file where values are separated by commas.
 You can open it with Notepad. The function is read.csv()

 "input/acids.csv" means: inside the folder I am standing
 in, go into input and take acids.csv. This is better than
 writing the full path, because then the script works on any
 computer.

```{r}
acids     <- read.csv("input/acids.csv")
insulin   <- read.csv("input/insulin.csv")
abundance <- read.csv("input/abundance.csv")
```

### input/acids.csv
The six organic acids measured in each sample, in ug/ml.
One row per sample.

```{r}
acids
str(acids)
dim(acids)
names(acids)
summary(acids)
```

```{r}
table(acids$product)
sum(is.na(acids))       # any missing values? (should be 0)
```

### input/insulin.csv
Insulin released by the cells, in ng per million cells and
hour. 6 samples x 3 experiments = 18 rows.

```{r}
head(insulin)
nrow(insulin)
table(insulin$product, insulin$experiment)
```

### input/abundance.csv
Which bacteria were in each sample, as a % of the total.
All 190 species detected in the study, one per row.

```{r}
dim(abundance)
head(abundance)
```

CAREFUL: 190 rows. Typing  abundance  on its own floods the
console. With big tables you look at them like this instead.

Two things worth knowing about this table:

1) The columns do NOT add up to exactly 100:

```{r}
colSums(abundance[, -1])
```

   Why? To find out which bacteria are in a sample, its DNA is
   sequenced and the pieces are compared against a database of
   known species. Some pieces do not match anything and stay
   UNCLASSIFIED, so they cannot appear in the table: we do not
   know which species they belong to. Each column therefore
   adds up to  100 - (% of unclassified reads), which in this
   study ranges from 0% to 12.3%. It is not an error.

   Note the  [, -1]  : it drops the first column (species),
   which is text. colSums() cannot add up words.

2) Some species are 0 in all six samples. They were in the
   original study, which had 26 samples, but not in ours.

```{r}
sum(rowSums(abundance[, -1]) == 0)
```

### input/taxonomy.csv
The record card of each species: kingdom, phylum, class,
order, family and genus. 190 rows, in the same order as
abundance.csv, so the two tables can be matched up.

```{r}
taxonomy <- read.csv("input/taxonomy.csv")
head(taxonomy)
dim(taxonomy)
```

How many species belong to each phylum

```{r}
table(taxonomy$phylum)
```

All the species of the genus Bifidobacterium

```{r}
taxonomy[taxonomy$genus == "Bifidobacterium", ]
```

### EXERCISE 5
Using what you learned in the first part (no dplyr yet):
  a) pull out the butyric column of  acids
  b) compute the mean butyric acid of the 6 samples
  c) keep only the rows of product "B"

  you need:  $   mean()   [ , ]   ==

> your answer:

```{r}

```

### EXERCISE 6
Which sample has the MOST butyric acid?
Hint: use max() and then a condition.

  you need:  max()   ==   [ , ]

> your answer:

```{r}

```

## 12. SAVING RESULTS: THE output FOLDER

   Golden rule: the input files are NEVER modified.
   Everything R produces goes into output.

Example: we save a small table

```{r}
only_B <- acids[acids$product == "B", ]
write.csv(only_B, "output/product_B_only.csv", row.names = FALSE)
```

```{r}
list.files("output")
```

row.names = FALSE stops R from adding a column with 1,2,3...

### EXERCISE 7
Save a file called  acids_control.csv  in output, containing
only the Control rows.

  you need:  $   [ , ]   ==   write.csv()   list.files()

> your answer:

```{r}

```

## END OF SESSION 1

 What to take away:
   <-           store something in a variable
   c()          build a vector
   [ ]          positions
   class()      what type of data it is
   table()      count categories
   data.frame   the table
   $            one column
   str/nrow/head/summary   LOOK at the data before analysing
   read.csv("input/file.csv")     read data
   write.csv(..., "output/...")   save results
   Never touch the input files
