---
title: "Basic R Course - Session 2 (Handling tables with dplyr)"

---


```{r}
# install.packages("dplyr")    # only the first time

library(dplyr)

acids <- read.csv("input/acids.csv")

acids
dim(acids)     # 6 rows (the samples) and 9 columns
```

# 1. The 6 dplyr verbs

dplyr is a package for handling tables. It is useful for the things you do with
data every day: keeping only the rows you want, choosing columns, sorting,
creating new columns, and calculating summaries per group. You could already do
most of this with base R in session 1, but dplyr makes it simpler and easier to
read.

- `filter()` picks ROWS
- `select()` picks COLUMNS
- `arrange()` sorts the rows
- `mutate()` adds a new column
- `summarise()` reduces to a single number
- `group_by()` does the above per group

They are all written the same way, and the first argument is ALWAYS the table: `verb(table, what_I_want)`

## filter() -> picks ROWS

You give it a condition and it keeps the rows that meet it. The columns are untouched.

```{r}
# "give me the rows whose product is B"
filter(acids, product == "B")
#   from 6 rows -> 2 remain

# "give me the rows with more than 1000 ug/ml of butyric acid"
filter(acids, butyric > 1000)
#   from 6 rows -> 3 remain
```

CAREFUL: to compare you write a DOUBLE equals sign `==`

Column names go WITHOUT quotes (`product` is a column). Text values go WITH
quotes (`"B"` is a value inside that column). Numbers go without quotes.
Worth triggering both errors on purpose in class: a single `=` gives "We
detected a named input", and `product == B` gives "object 'B' not found".

### Exercise 1a

Keep only the Control rows.

```{r}

```


### Exercise 1b

Now a filter with a number instead of a word: keep only the rows whose acetic acid is above 3000.

```{r}

```


## select() -> picks COLUMNS

The opposite of `filter()`. All rows are kept.

```{r}
# "give me only these two columns"
select(acids, sample, butyric)
#   still 6 rows, but out of 9 columns only 2 remain

# A minus sign REMOVES a column
select(acids, -formic)

# A colon gives a RANGE of columns
select(acids, formic:citric)
```

### Exercise 2a

Keep only the columns sample, product and acetic.

```{r}

```


### Exercise 2b

Now the other way round: show every column EXCEPT propionic (use the minus sign).

```{r}

```


## arrange() -> sorts the rows

It removes nothing and adds nothing, it only changes order.

```{r}
# sort the samples by butyric acid
arrange(acids, butyric)          # from lowest to highest
arrange(acids, desc(butyric))    # desc = highest to lowest
```

### Exercise 3a

Sort the samples from highest to lowest propionic acid.

```{r}

```


### Exercise 3b

Now without `desc()`: sort the samples from lowest to highest formic acid.

```{r}

```


## mutate() -> adds a new column

Computes something from the existing columns and sticks it on at the end.

The "good" fatty acids are butyric and propionic:

```{r}
# add the two "good" acids into a single new column
mutate(acids, scfa = butyric + propionic)
#   an scfa column appears at the end: 9 columns -> 10
#   (scfa = short chain fatty acids)
```

IMPORTANT: `mutate()` only shows it on screen. To keep it in the table you have to store it with `<-`:

```{r}
acids <- mutate(acids, scfa = butyric + propionic)
acids
```

### Exercise 4a

Create a column `ratio` that is butyric divided by acetic, rounded to 2 decimals.

```{r}

```


### Exercise 4b

Now an easier one, just a sum: create a column `acetic_lactic` that adds acetic and lactic.

```{r}

```


### Exercise 4.1 (quick review: arrange + select)

No `group_by()` here, just a mix of two verbs you already know, using the `scfa`
column you created with `mutate()`. Two separate lines:

a) sort the samples from highest to lowest scfa
b) show only the columns sample and scfa

```{r}

```


## summarise() -> reduces to a single number

It condenses all the rows into one.

In session 1 we would have written this:

```{r}
mean(acids$butyric)
```

`summarise()` does exactly the same thing, but hands it back as a TABLE, with a column name that we choose:

```{r}
# the average butyric acid across all 6 samples, as a table
summarise(acids, mean_butyric = mean(butyric))
#   from 6 rows -> 1 row
```

So why bother, if `mean()` already worked? Because of what comes next: `summarise()` can do this per group, and plain `mean()` cannot.

You can ask for several numbers at once:

```{r}
summarise(acids,
          mean_butyric   = mean(butyric),
          mean_propionic = mean(propionic))
```

## group_by() -> doing it per group

Everything above gives ONE number for the whole table. But what we actually want is one number for the Control, one for A and one for B, so that we can compare them.

**First, how we would do it BY HAND.** Pick out one product, compute its mean. Then the next one.

```{r}
control_rows <- filter(acids, product == "Control")
mean(control_rows$butyric)          # 491.8

a_rows <- filter(acids, product == "A")
mean(a_rows$butyric)                # 1237.5

b_rows <- filter(acids, product == "B")
mean(b_rows$butyric)                # 2319.8
```

It works. But it took 6 lines for 3 products, the results come out loose instead of in a table, and with 20 products it would be unbearable.

**Now the same thing with group_by():**

```{r}
# average butyric acid per product, in one table
by_product <- group_by(acids, product)
summarise(by_product, mean_butyric = mean(butyric))
```

Same three numbers, in one table, in two lines.

`group_by()` on its own displays nothing new: all it does is mark which rows belong together. The work is done by the `summarise()` that comes after it, which now runs once per group instead of once for the whole table.

### Exercise 6a

Now the same thing, but with PROPIONIC acid: compute its mean per product. Two lines: first group, then summarise.

```{r}

```


### Exercise 6b

A different question: how MANY samples does each product have? Use `n()`, which counts the rows in each group.

```{r}

```


# 2. Chaining steps

One verb is almost never enough. To do several things in a row you store each step in a variable with `<-` and pass it to the next one. It is like an assembly line.

```{r}
# step 1: keep only replicate 1's rows
r1_only <- filter(acids, replicate == 1)   # 6 rows -> 3
# step 2: from that result, keep three columns
small   <- select(r1_only, sample, formic, acetic)
small
```

The result of step 1 is the input table of step 2. Give them names that say what they contain, not step1/step2.

## The pipe (optional, just so you recognise it)

On the internet you will constantly see the symbol `|>` (or the older `%>%`), the "pipe". It chains the same steps without inventing a name for each one: it takes the table on the left and hands it to the next verb as its first argument.

```{r}
# exactly the same as the two lines above, but in one line
acids |> filter(replicate == 1) |> select(sample, formic, acetic)
```

You do NOT need the pipe for this course. We show it once, actually used, so that it does not startle you when you meet it online.

### Exercise 8

Keep the rows of product B and sort them from highest to lowest butyric acid, storing each step in a variable.

```{r}

```


### Exercise 8b (optional: with the pipe)

Redo Exercise 8 in a single line using the pipe `|>`, without storing anything in between.

```{r}

```


# 3. Saving your results

Anything you can see, you can save. Same as session 1: results go to `output/`,
and `input/` is never touched.

```{r}
by_product <- group_by(acids, product)
summary_table <- summarise(by_product,
                           mean_butyric   = mean(butyric),
                           mean_propionic = mean(propionic))
summary_table

write.csv(summary_table, "output/acids_summary.csv", row.names = FALSE)

list.files("output")   # check it is really there
```

`row.names = FALSE` stops R from adding a useless first column of row numbers.

### Exercise 9

Chaining two verbs, storing each step:

a) keep the rows of replicate 2
b) sort them from highest to lowest propionic acid
c) save the result in `output/replicate2_propionic.csv`

```{r}

```


### Exercise 10 (everything at once)

a) add a column `scfa` with butyric plus propionic
b) group by product and calculate the mean `scfa` of each one
c) sort the result from highest to lowest
d) save it in `output/scfa_per_product.csv`

```{r}

```


### Exercise 11 (two numbers per group)

a) for each product, calculate the mean butyric acid AND the highest butyric acid
b) which product has the biggest gap between the two? What does that tell you?

```{r}

```


> b) Product A: its mean and its maximum are 400 apart, while in B they almost
> coincide. Without naming it yet, they are already seeing that A's two samples
> disagree. Good bridge to the next section: "keep that in mind, we come back to
> it in ten minutes".

### Exercise 12 (filter first, then count)

a) using the `scfa` column, keep only the samples whose `scfa` is above 1000
b) count how many samples of each product survived

```{r}

```


> The Control comes out with a 1: one of its two samples passes the threshold and
> the other does not. Another hint of what is coming.

# 4. Do the two replicates agree?

Each condition was fermented twice, separately (R1 and R2). Before summarising anything you have to check whether the two give the same answer. If they do, the experiment is reproducible.

**Biological replicate vs technical repetition.** A biological replicate is a real
repetition of the experiment: two separate fermentation wells, inoculated
separately and left to ferment on their own. That is the `replicate` column, R1
and R2. A technical repetition is measuring the same material more than once:
the `experiment` column of the insulin table. It tells you how good your
technique is, not how repeatable the biology is. Technical repetitions get
averaged into a single value; they never increase your sample size. Treating
them as independent samples inflates n and hands you significance that is not
real.

We will compare ALL SIX acids, not just one. If we only looked at one, the answer would depend on which one we happened to pick.

**Step 1:** we split the two replicates into two tables.

```{r}
r1 <- filter(acids, replicate == 1)     # Control_R1, A_R1, B_R1
r2 <- filter(acids, replicate == 2)     # Control_R2, A_R2, B_R2

r1
r2
```

Same products, same columns, same order in both. That is what makes the rest work.

**Step 2:** we keep only the six acid columns of each table.

```{r}
values_r1 <- select(r1, formic:citric)
values_r2 <- select(r2, formic:citric)

values_r1
values_r2
```

**Step 3:** we compare them, one idea per line.

When you subtract two tables of the same size, R subtracts them cell by cell: the formic of Control_R1 minus the formic of Control_R2, and so on for all 18 numbers.

```{r}
gap <- abs(values_r1 - values_r2)     # HOW FAR APART the two are
gap
```

Why `abs()`? Without it we would get negative numbers whenever R2 is the bigger one. The minus sign only tells us WHICH one was bigger, and we do not care about that: we want to know how far apart they are. `abs()` drops the sign and keeps the size. Try it without `abs()` and compare.

```{r}
average <- (values_r1 + values_r2) / 2   # the TYPICAL VALUE there
average
```

CAREFUL: this is not the mean of an acid across samples. It is the middle point between the two replicates of that one acid in that one condition.

```{r}
diff_pct <- round(100 * gap / average, 1)   # gap, as a % of the typical value
diff_pct
```

**WHY DO WE DIVIDE?** Two real examples from this table:

```
                     R1       R2      gap   typical value    %
  butyric in A     1636      839      796         1238      64
  formic  in B    22334    26137     3804        24236      16
```

If we stopped at the gap, the formic acid of B (3804) would look FIVE TIMES WORSE than the butyric acid of A (796). It is the other way round: 3804 out of 24000 is small change, while 796 out of 1238 is enormous.

Two people weigh a bag of flour: 1000 g and 1010 g, you would say they agree. Two
people weigh a letter: 20 g and 30 g. Same ten grams, completely different
meaning. The only way to compare them is to ask: ten out of how much?

Dividing also makes the six acids comparable with each other, which is what lets us read the whole table at once.

And we glue the product names back on, to read it easily:

```{r}
comparison <- data.frame(product = r1$product, diff_pct)
comparison
```

```
product formic acetic propionic lactic butyric citric
Control    2.7    4.9      60.1    4.6    37.9    2.9
      A    0.7   11.7      43.6   29.2    64.3   32.9
      B   15.7    6.7      25.1    3.7     5.5   14.6
```

## Reading the table (discussion, no code)

Remember what each number means: "the two replicates differ by this percentage of the normal value". Small = they agree.

a) Look at the three rows. Which product has the most reproducible replicates, that is, the smallest numbers?
b) Now look at the six columns. Which TWO acids give the biggest numbers in all three products?

# 5. The same thing as a picture

This is how the original paper presented it: a bar for the mean, an error bar for
the standard deviation, and the individual replicates as dots.

```{r}
library(tidyr)
library(ggplot2)

# order of the products in the plot (otherwise R sorts them alphabetically)
acids$product <- factor(acids$product, levels = c("Control", "A", "B"))

# long format: one row per sample and acid
acids_long <- pivot_longer(acids, cols = formic:citric,
                           names_to = "acid", values_to = "concentration")

# mean and SD per product and acid
by_group      <- group_by(acids_long, product, acid)
acids_summary <- summarise(by_group,
                           mean = mean(concentration),
                           sd   = sd(concentration))

ggplot() +
  geom_col(data = acids_summary, aes(x = product, y = mean), fill = "orange") +
  geom_errorbar(data = acids_summary,
                aes(x = product, ymin = mean - sd, ymax = mean + sd), width = 0.2) +
  geom_point(data = acids_long, aes(x = product, y = concentration)) +
  facet_wrap(~ acid, scales = "free_y") +
  labs(x = NULL, y = "Concentration (ug/ml)")
```

With two replicates the standard deviation is just the gap divided by the square
root of 2, so the plot and the table of section 4 carry the same information.

**Butyric acid, the panel to comment on.** B's two dots sit almost on top of each
other (tiny error bar): its replicates agree. A's dots are far apart (long error
bar): they do not. Careful with the two different overlaps here: two dots close
together within one product is good (the replicates agree), whereas the error
bars of two different products overlapping is bad (you cannot tell those products
apart). And they are connected: A's replicates disagree so much that its error
bar stretches down towards the Control, so A is no longer clearly different from
it. B sits far above the Control, much further than its own replicates move, so
that difference survives.

**CONCLUSIONS:**

1. Product B is the most reproducible condition, product A the least. Now we can say this looking at a whole row, not at a single number.
2. Formic, acetic and lactic acid repeat almost exactly. Propionic and butyric are the noisy ones, in every condition.
3. And here is the uncomfortable part: propionic and butyric are precisely the two acids that matter for our hypothesis, the ones linked to insulin. The measurements we care about most are the least reproducible ones. That is why the authors of the original paper report all these acids descriptively, without statistics: with two wells there is not much else you can honestly do.
4. What we do with that: we never delete the replicate we do not like, and we ask whether the difference between products is bigger than the disagreement within a product. For butyric acid that works for B, but not for A.

That comparison, signal against noise, is exactly what a statistical test does for you. You have just done it by eye.

The lesson: reproducibility is not the same in every condition or for every measurement, and you have to LOOK at it rather than assume it.

---

# END OF SESSION 2 (dplyr)

- `filter()` picks ROWS (trims vertically)
- `select()` picks COLUMNS (trims horizontally)
- `arrange()` sorts
- `mutate()` adds a column
- `summarise()` reduces
- `group_by()` makes summarise work per group

Chain steps by storing each result in a variable. Check the replicates BEFORE summarising them.

(You have also met the pipe `|>` once, in section 2. You do not need it for this course; we just wanted it not to startle you when you see it online.)
